# Dissertation part 2: Data analysis

Part 2 contains the primary analysis portion of my dissertation. The aim of this analysis is to assert which factors impact terrorist attack occurrence using logistic regression, before combining these factors into a stack and employing the Maximum Entropy Model (MAXENT) to infer areas at risk of terrorist attacks given predictor variables.

```{r Loading packages}
library(sf)
library(tmap)
library(ggplot2)
library(dplyr)
library(lubridate)
library(raster)
library(extrafont)
library(spdep)
library(geostan)
library(SpatialEpi)
library(dismo)
library(rJava)
library(terra)
library(pscl)
library(exactextractr)
library(extrafont)
```

## Preparing factor data 

For my model I will be dealing with point occurrence data of terrorist attacks in Borno state from 2010 to 2020 as well as raster data for each predictor variable. 

```{r Loading factor data}
population <- raster('nga_ppp_2020_1km_Aggregated.tif')
male_literacy <- raster('NGA_literacy_M.tif')
female_literacy <- raster('NGA_literacy_F.tif')
school_children <-raster('NGA_F_M_SECONDARY_2020_1km.tif')
poverty <- raster('nga10povcons125.tif')
urban_rural <- raster('nga_bsgme_v0a_100m_2020.tif')
topography <- raster('nga_srtm_topo_100m.tif')
roads <- raster('nga_osm_dst_road_100m_2016.tif')
education_facilities <- raster('educational_facilities_raster.tif')
mosques <- raster('mosques_Raster.tif')
```

```{r Clipping, reprojecting and resampling my rasters}
# Preparing my extents
borno_extent <- extent(borno)
pcs <- CRS("+proj=longlat +ellps=WGS84 +datum=WGS84 +no_defs+ towgs84=0,0,0")

# Function to clip, mask, and project a raster
process_raster <- function(raster, extent, outline, crs) {
  clipped <- crop(raster, extent)
  masked <- mask(clipped, outline)
  projected <- projectRaster(masked, crs=crs)
  return(projected)
}

# Apply the function to each raster variable
population_processed <- process_raster(population, borno_extent, borno, pcs)
male_literacy_processed <- process_raster(male_literacy, borno_extent, borno, pcs)
female_literacy_processed <- process_raster(female_literacy, borno_extent, borno, pcs)
school_children_processed <- process_raster(school_children, borno_extent, borno, pcs)
poverty_processed <- process_raster(poverty, borno_extent, borno, pcs)
urban_rural_processed <- process_raster(urban_rural, borno_extent, borno, pcs)
topography_processed <- process_raster(topography, borno_extent, borno, pcs)
roads_processed <- process_raster(roads, borno_extent, borno, pcs)
education_facilities_processed <- process_raster(education_facilities, borno_extent, borno, pcs)
mosques_processed <- process_raster(mosques, borno_extent, borno, pcs)

# Standardize all rasters to dimensions of approx. 2.5 km (area = 21 sq.km)
borno_extent <- extent(borno)  
RasterTemplate <- raster(crs = pcs, res = c(2.5/111.32, 2.5/111.32))
extent(RasterTemplate) <- borno_extent

# Resample the rasters
population_resampled <- resample(population_processed, RasterTemplate, method = "bilinear")
male_literacy_resampled <- resample(male_literacy_processed, RasterTemplate, method = "bilinear")
female_literacy_resampled <- resample(female_literacy_processed, RasterTemplate, method = "bilinear")
school_children_resampled <- resample(school_children_processed, RasterTemplate, method = "bilinear")
poverty_resampled <- resample(poverty_processed, RasterTemplate, method = "bilinear")
urban_rural_resampled <- resample(urban_rural_processed, RasterTemplate, method = "ngb")
topography_resampled <- resample(topography_processed, RasterTemplate, method = "bilinear")
roads_resampled <- resample(roads_processed, RasterTemplate, method = "bilinear")
education_facilities_resampled <- resample(education_facilities_processed, RasterTemplate, method = "ngb")
mosques_resampled <- resample(mosques_processed, RasterTemplate, method = "ngb")

# Ensuring the binary of resampled rasters remains
urban_rural_resampled <- calc(urban_rural_resampled, fun = function(x) ifelse(x >= 0.5, 1, 0))
education_facilities_resampled <- calc(education_facilities_resampled, fun = function(x) ifelse(x >= 0.5, 1, 0))
mosques_resampled <- calc(mosques_resampled, fun = function(x) ifelse(x >= 0.5, 1, 0))

# Assign names to the resampled rasters
names(population_resampled) <- "population"
names(male_literacy_resampled) <- "male_literacy"
names(female_literacy_resampled) <- "female_literacy"
names(school_children_resampled) <- "school_children"
names(poverty_resampled) <- "poverty"
names(urban_rural_resampled) <- "urban_rural"
names(topography_resampled) <- "topography"
names(roads_resampled) <- "roads"
names(education_facilities_resampled) <- "education_facilities"
names(mosques_resampled) <- "mosques"

# Saving the rasters
writeRaster(population_resampled, filename = "population_raster_final.tif", overwrite = TRUE)
writeRaster(male_literacy_resampled, filename = "male_literacy_raster_final.tif", overwrite = TRUE)
writeRaster(female_literacy_resampled, filename = "female_literacy_raster_final.tif", overwrite = TRUE)
writeRaster(school_children_resampled, filename = "school_children_raster_final.tif", overwrite = TRUE)
writeRaster(poverty_resampled, filename = "poverty_raster_final.tif", overwrite = TRUE)
writeRaster(urban_rural_resampled, filename = "urban_rural_raster_final.tif", overwrite = TRUE)
writeRaster(topography_resampled, filename = "topography_raster_final.tif", overwrite = TRUE)
writeRaster(roads_resampled, filename = "roads_raster_final.tif", overwrite = TRUE)
writeRaster(education_facilities_resampled, filename = "education_raster_final.tif", overwrite = TRUE)
writeRaster(mosques_resampled, filename = "mosques_raster_final.tif", overwrite = TRUE)
```

```{r Re-reading finished rasters}
population_resampled <- raster('population_raster_final.tif')
male_literacy_resampled <- raster('male_literacy_raster_final.tif')
female_literacy_resampled <- raster('female_literacy_raster_final.tif')
school_children_resampled <-raster('school_children_raster_final.tif')
poverty_resampled <- raster('poverty_raster_final.tif')
urban_rural_resampled <- raster('urban_rural_raster_final.tif')
topography_resampled <- raster('topography_raster_final.tif')
roads_resampled <- raster('roads_raster_final.tif')
education_facilities_resampled <- raster('education_raster_final.tif')
mosques_resampled <- raster('mosques_raster_final.tif')
```

Now that all my predictor variables are in the correct crs and spatial extent (clipped to Borno shapefile), I will quickly visualise these. 

```{r Plotting predictor variable rasters}
# Population
population_map <- tm_shape(population_resampled) +
  tm_raster(style = "cont", title = "Population /2.5km2", palette = "-RdYlBu") +
  tm_shape(borno) +
  tm_polygons(alpha = 0, border.col = "black") +
  tm_shape(borno_LGAs) +  
  tm_borders(col = "black", lwd = 0.3) + 
  tm_layout(frame = FALSE, legend.position = c("left", "top"),
            title.fontfamily = "Avenir", 
            legend.text.fontfamily = "Avenir", 
            legend.title.fontfamily = "Avenir")

# Male literacy
male_lit_map <- tm_shape(male_literacy_resampled) + 
  tm_raster(style = "cont", title = "%", palette= "Oranges", breaks = seq(0, 1, by = 0.2), labels = c("0", "20", "40", "60", "80", "100")) +
  tm_shape(borno) + 
  tm_polygons(alpha = 0, border.col = "black") +
  tm_shape(borno_LGAs) + 
  tm_borders(col = "black", lwd = 0.3) +
  tm_layout(frame = FALSE, legend.position = c("left", "top"),
            title.fontfamily = "Avenir", 
            legend.text.fontfamily = "Avenir", 
            legend.title.fontfamily = "Avenir")

# Female literacy
female_lit_map <- tm_shape(female_literacy_resampled) +
  tm_raster(style = "cont", title = "%", palette = "Blues", breaks = seq(0, 1, by = 0.2), labels = c("0", "20", "40", "60", "80", "100")) +
  tm_shape(borno) +
  tm_polygons(alpha = 0, border.col = "black") +
  tm_shape(borno_LGAs) + 
  tm_borders(col = "black", lwd = 0.3) +
  tm_layout(frame = FALSE, legend.position = c("left", "top"),            
            title.fontfamily = "Avenir", 
            legend.text.fontfamily = "Avenir", 
            legend.title.fontfamily = "Avenir")

# School children
school_children_map <- tm_shape(school_children_resampled) +
  tm_raster(style = "cont", title = "/2.5km2", palette = "-RdYlBu") +
  tm_shape(borno) +
  tm_polygons(alpha = 0, border.col = "black") +
  tm_shape(borno_LGAs) + 
  tm_borders(col = "black", lwd = 0.3) +
  tm_layout(frame = FALSE, legend.position = c("left", "top"),
            title.fontfamily = "Avenir", 
            legend.text.fontfamily = "Avenir", 
            legend.title.fontfamily = "Avenir")

# Poverty
poverty_map <- tm_shape(poverty_resampled) +
  tm_raster(style = "cont", title = "% living on >$1.25/d", palette = "-RdYlBu", breaks = seq(0, 1, by = 0.2), labels = c("0", "20", "40", "60", "80", "100")) +
  tm_shape(borno) +
  tm_polygons(alpha = 0, border.col = "black") +
  tm_shape(borno_LGAs) + 
  tm_borders(col = "black", lwd = 0.3) +
  tm_layout(frame = FALSE, legend.position = c("left", "top"),
            title.fontfamily = "Avenir", 
            legend.text.fontfamily = "Avenir", 
            legend.title.fontfamily = "Avenir")

# Urban v rural
urban_rural_map <- tm_shape(urban_rural_resampled) +
  tm_raster(style = "cat", title = "Land Type", palette = c("green", "blue"), labels = c("Rural", "Urban")) +
  tm_shape(borno) +
  tm_polygons(alpha = 0, border.col = "black") +
  tm_shape(borno_LGAs) + 
  tm_borders(col = "black", lwd = 0.3) +
  tm_layout(frame = FALSE, legend.position = c("left", "top"),
            title.fontfamily = "Avenir", 
            legend.text.fontfamily = "Avenir", 
            legend.title.fontfamily = "Avenir")

# Topography
topography_map <- tm_shape(topography_resampled) +
  tm_raster(style = "cont", title = "Elevation (m)", palette= "Greys") +
  tm_shape(borno) +
  tm_polygons(alpha = 0, border.col = "black") +
  tm_shape(borno_LGAs) + 
  tm_borders(col = "black", lwd = 0.3) +
  tm_layout(frame = FALSE, legend.position = c("left", "top"),
            title.fontfamily = "Avenir", 
            legend.text.fontfamily = "Avenir", 
            legend.title.fontfamily = "Avenir")

# Roads
road_map <- tm_shape(roads_resampled) +
  tm_raster(style = "cont", title = "Distance (km)", palette= "Oranges") +
  tm_shape(borno) +
  tm_polygons(alpha = 0, border.col = "black") +
  tm_shape(borno_LGAs) + 
  tm_borders(col = "black", lwd = 0.3) +
  tm_layout(frame = FALSE, legend.position = c("left", "top"),
            title.fontfamily = "Avenir", 
            legend.text.fontfamily = "Avenir", 
            legend.title.fontfamily = "Avenir")

# Educational facilities
education_facilities_map <- tm_shape(education_facilities_resampled) +
  tm_raster(style = "cat", title = "Facility presence", palette = c("yellow", "white"), labels = c("Facilities", "No facilities")) +
  tm_shape(borno) +
  tm_polygons(alpha = 0, border.col = "black") +
  tm_shape(borno_LGAs) + 
  tm_borders(col = "black", lwd = 0.3) +
  tm_layout(frame = FALSE, legend.position = c("left", "top"),
            title.fontfamily = "Avenir", 
            legend.text.fontfamily = "Avenir", 
            legend.title.fontfamily = "Avenir")

# Mosques
mosques_map <- tm_shape(mosques_resampled) +
  tm_raster(style = "cat", title = "Mosque presence", palette = c("purple", "white"), labels = c("Mosques", "No mosques")) +
  tm_shape(borno) +
  tm_polygons(alpha = 0, border.col = "black") +
  tm_shape(borno_LGAs) + 
  tm_borders(col = "black", lwd = 0.3) +
  tm_layout(frame = FALSE, legend.position = c("left", "top"),
            title.fontfamily = "Avenir", 
            legend.text.fontfamily = "Avenir", 
            legend.title.fontfamily = "Avenir")
```

```{r Plotting rasters}
# Plotting variables
population_map
male_lit_map
female_lit_map
school_children_map
poverty_map
urban_rural_map
topography_map
road_map
education_facilities_map
mosques_map

# Downloading raster plots
tmap_save(population_map, filename = "population_map.png", width = 2000, height = 2000, dpi = 300)
tmap_save(male_lit_map, filename = "male_lit_map.png", width = 2000, height = 2000, dpi = 300)
tmap_save(female_lit_map, filename = "female_lit_map.png", width = 2000, height = 2000, dpi = 300)
tmap_save(school_children_map, filename = "school_children_map.png", width = 2000, height = 2000, dpi = 300)
tmap_save(poverty_map, filename = "poverty_map.png", width = 2000, height = 2000, dpi = 300)
tmap_save(urban_rural_map, filename = "urban_rural_map.png", width = 2000, height = 2000, dpi = 300)
tmap_save(topography_map, filename = "topography_map.png", width = 2000, height = 2000, dpi = 300)
tmap_save(road_map, filename = "road_map.png", width = 2000, height = 2000, dpi = 300)
tmap_save(education_facilities_map, filename = "education_facilities_map.png", width = 2000, height = 2000, dpi = 300)
tmap_save(mosques_map, filename = "mosques_map.png", width = 2000, height = 2000, dpi = 300)
```

Now I am ready to proceed with logistic regression and MAXENT analysis

## Logistic regression 

```{r Logistic regression}
# Creating a multi-band raster
envCovariates <- stack(population_resampled, male_literacy_resampled, female_literacy_resampled, school_children_resampled, poverty_resampled, urban_rural_resampled, topography_resampled, roads_resampled, education_facilities_resampled, mosques_resampled)
names(envCovariates) <- c("Population", "Male Literacy", "Female Literacy", "School children", "Poverty", "Urban v Rural", "Topography", "Roads", "Educational facilities", "Mosques")

# Turning my terrorism point dataset into an sp object - essential for the code
borno_terrorism_sp <- as(borno_terrorism, "Spatial")

# Splitting the data into training and testing sets using k-fold
set.seed(20000430)
group <- kfold(borno_terrorism_sp, 5)
occ_train <- borno_terrorism_sp[group != 1, ]
occ_test <- borno_terrorism_sp[group == 1, ]

# Extracting the predictor variables for training and testing
train_vals <- raster::extract(envCovariates, occ_train)
test_vals <- raster::extract(envCovariates, occ_test)

# Creating background points
set.seed(20000430)
bg_points <- as.data.frame(randomPoints(envCovariates, 2 * nrow(occ_train)))
coordinates(bg_points) <- ~ x + y

# Extracting predictor variables for background points
bg_vals <- raster::extract(envCovariates, bg_points)

# Combining the presence and absence data
pres_abs <- c(rep(1, nrow(train_vals)), rep(0, nrow(bg_vals)))
data_combined <- as.data.frame(rbind(train_vals, bg_vals))
logistic_data <- as.data.frame(cbind(pres_abs, data_combined))
colnames(logistic_data)[1] <- "terror_attack"

# Checking the column names of my logistic_data dataset
print(colnames(logistic_data))

# Running the Logistic regression model
log_model <- glm(terror_attack ~ Population + Male.Literacy + Female.Literacy + School.children + Poverty + Urban.v.Rural + Topography + Roads + Educational.facilities + Mosques,
                 family = binomial(link = "logit"), data = logistic_data)
```

To assess my model outputs:

```{r Assessing outputs 1}
# To examine the P-values
summary(log_model) 
```

Statistically significant predictors of terrorist attacks are population, male literacy, school children, poverty, urban v rural, roads, educational facilities and mosques.

Factors with a positive relationship to terrorist attacks (ie. as the variables increase, the likelihood of an attack increases) are population, male literacy and urban v rural.

Factors with a negative relationship to terrorist attacks (ie. as the variables increase, the likelihood of an attack decreases) are female literacy, school children, poverty, roads, educational facilities and mosques.

Topography is not a significant predictor.

Approximately 50.66% of variance is explained by the model - a good fit for LR models.

Broadly, the model indicates that all of my selected factors bar topography influence the likelihood of terrorist attacks.

Next I will look at the odds ratios, as these enable me to understand the effect size of predictors in my model. I must exponentiate my coefficients for this.

```{r Assessing outputs 2}
# Converting to odds ratios
exp(coefficients(log_model))

# Extracting McFadden's R2
pR2(log_model)
```
One unit increase in population increases the odds of an attack by approx. 0.9%.

One unit increase in male literacy increases the odds of an attack by approx. 781%

One unit increase in female literacy decreases the odds of an attack by approx. 66%

One unit increase in school children decreases the odds of an attack by approx. 5.9%

One unit increase in poverty decreases the odds of an attack by approx. 92.3%

One unit increase in urban v rural increases the odds of an attack by approx. 1819%

One unit increase in topography slightly increases the odds of an attack by approx. 0.07%

One unit increase in roads decreases the odds of an attack by approx. 14.2%

One unit increase in educational facilities decreases the odds of an attack by approx. 36.9%

One unit increase in mosques decreases the odds of an attack by approx. 31%

The pseudo-R2 values indicate the model has a good fit and explains a substantial portion of the variance in the data.

Thus broadly, the model performs well, with several significant predictors and good fit statistics. This means that for my MAXENT model I will include all but topography as this is not statistically significant

## MAXENT analysis

To employ my MAXENT model I must first prepare my data. This involves creating pseudo-background points as absence. Background data ensures I'm not attempting to guess point locations where an event is absent. Rather, I'm trying to characterise the environment of the study region. Background data establishes the environmental domain of the study, whilst presence data will establish under which conditions a terrorist attack is more likely to occur than on average. Essentially, these are control points, generated at random. 

```{r MAXENT (i)}
# Turning my terrorism point dataset into an sp object - essential for the code
borno_terrorism_sp <- as(borno_terrorism, "Spatial")

# Redefining my variables to remove topography given its statistical insignificance indicated by the logistic regression
variables <- stack(population_resampled, male_literacy_resampled, female_literacy_resampled, school_children_resampled, poverty_resampled, urban_rural_resampled, roads_resampled, education_facilities_resampled, mosques_resampled)
names(variables) <- c("Population", "Male Literacy", "Female Literacy", "School children", "Poverty", "Urban v Rural", "Roads", "Educational facilities", "Mosques")

# Setting the seed
set.seed(20000430)

# Using spsample() to generate twice number of terrorist attack points randomly within Borno's border
background_points <- spsample(borno_sp, n=2*length(borno_terrorism_sp), "random")
```

Now I will extract all the raster values from predictor variables onto presence-absence points. This will produce matricies rather than data frames.

```{r MAXENT (ii)}
# Performing raster extraction from the predictor covariates on to points
terrorist_attacks_env <- extract(variables, borno_terrorism_sp)
background_points_env <- extract(variables, background_points)
```

For presence points, an indicator of 1 must be added to signify presence. For background points, the opposite (0 for absence). This is done because I am modelling a probability and niche models take outcomes that are either a Bernoulli or Binomial distribution.

```{r MAXENT (iii)} 
# Converting the large matrix objects to data frame objects and add outcome `attack` indicator
terrorist_attacks_env <-data.frame(terrorist_attacks_env,attack=1)
background_points_env <-data.frame(background_points_env,attack=0)

# View one of the data frame
head(terrorist_attacks_env, n=5)
head(background_points_env, n=5)
```

Final step is setting up the K-fold cross-validation. 

```{r MAXENT (iiii)}
# Splitting the plot panel into 4 segments for 4 AUC plots
par(mfrow=c(2,2))

# Creating a list() object to dump results inside `eMAX`
eMAX<-list()

folds <- 4

kfold_pres <- kfold(terrorist_attacks_env, folds)
kfold_back <- kfold(background_points_env, folds)

set.seed(20000430)

# Adapted loop code from https://rpubs.com/mlibxmda/GEOG70922_Week5
for (i in 1:folds) {
    train <- terrorist_attacks_env[kfold_pres!= i,]
    test <- terrorist_attacks_env[kfold_pres == i,]
    backTrain<-background_points_env[kfold_back!=i,]
    backTest<-background_points_env[kfold_back==i,]
    dataTrain<-rbind(train,backTrain)
    dataTest<-rbind(test,backTest)
    maxnet_eval <- maxent(x=dataTrain[,c(1:9)], p=dataTrain[,10], args=c("responsecurves"))
    eMAX[[i]] <- evaluate(p=dataTest[dataTest$attack==1,],a=dataTest[dataTest$attack==0,], maxnet_eval)
    plot(eMAX[[i]],'ROC')
}

aucMAX <- sapply( eMAX, function(x){slot(x, 'auc')} )
```

All the AUC values are above 0.9, which means the model has an exceptional discriminative ability (ie. can distinguish between attack versus absence points very well). Also, the fact that they're consistent suggests that the model is stable and performs well across different subsets of the data. The shape of the curves means that the model correctly identifies a high proportion of actual positives whilst making few false positive errors. 

```{r}
# Producing a report 4 of the AUC
aucMAX

# Finding the mean of AUC (and it must be > 0.50)
mean(aucMAX)

# Getting maxTPR+TNR for the model
Opt_MAX<-sapply( eMAX, function(x){ x@t[which.max(x@TPR + x@TNR)] } )
Opt_MAX

Mean_OptMAX<-mean(Opt_MAX)
Mean_OptMAX

# Combining all data
full_data <- rbind(terrorist_attacks_env, background_points_env)

final_maxent_model <- maxent(x = full_data[, c(1:9)], p = full_data[, 10], args = c("responsecurves"))

prob_attack <- predict(final_maxent_model, variables)
```

## Assessing model outputs

I want to look at which variable has the biggest contribution to the presence of terrorist attacks in Borno.

```{r Analysis: Predictor contribution}
plot(final_maxent_model, pch=19, xlab = "Percentage [%]", cex=1.2)
```
## Visualising attack risk

Next I will generate a predicted probability map from the above prob_attack object. 

```{r Mapping 1: Predicted probability}
tmap_mode("plot")

tm_shape(prob_attack) +
    tm_raster(title = "Predicted probability", palette = '-RdYlBu', style = 'cont', breaks = c(0, 0.2, 0.4, 0.6, 0.8, 1.0)) +
    tm_shape(borno) + 
    tm_polygons(alpha = 0, border.col = "black") +
    tm_layout(
      main.title = "Predicted probability of terrorist attacks", 
      title.size = 1, 
      legend.text.size = 0.4, 
      legend.title.size = 0.8, 
      fontfamily = "Avenir"
    )

tmap_save(predicted_prob, filename = "predicted_prob.png", width = 2000, height = 2000, dpi = 300)
```
Now I want to create a binary raster of expected trigger points for terrorist attacks across Borno. This will use the optimal threshold value from my cross validation object to reclassify the raster.

This will use the threshold value of 0.4377834, with anything estimated below this being classified as 0 whilst anything above as 1. 

```{r Mapping 2: Binary threshold}
# Reclassifying raster with threshold value 
create_classes_vector <- c(0, Mean_OptMAX, 0, Mean_OptMAX, 1, 1)
create_clasess_matrix <- matrix(create_classes_vector, ncol = 3, byrow = TRUE)
create_clasess_matrix

# Create new reclassify raster
suitability_attacks <- reclassify(prob_attack, create_clasess_matrix)

# Saving this
writeRaster(suitability_attacks, filename = "suitability_attacks.tif", overwrite = TRUE)

# Plot
tm_shape(suitability_attacks) + 
  tm_raster(style = "cat", title = "Threshold", palette = c("lightgrey", "#ef6030"), labels = c("Low threat", "Trigger points")) +
tm_shape(borno) + 
  tm_polygons(alpha = 0, border.col = "black") +
tm_layout(
  frame = FALSE, 
  legend.outside = TRUE, 
  fontfamily = "Avenir",
  title.color = "black",
  legend.text.color = "black"
) 
```

## Identifying at-risk populations

I want to now apply this predictive map to see which populations are most at risk of terrorist attacks. This is essential in informing resource allocation and safety of civilians. Firstly, I'm interested in looking at general population groups that are at risk, before quantifying under-15 populations at risk to support long-term resource allocation efforts.

Firstly, I will look at general populations at risk.

```{r Populations at high risk}
# Re-loading raster
suitability <- raster('suitability_attacks.tif')
population_resampled <- raster('population_raster_final.tif')
high_risk_population <- raster('high_risk_population.tif')

# Resampling the population raster to match the suitability raster
template <- raster(extent(suitability), res=res(suitability), crs=crs(suitability))
population_aligned <- resample(population_resampled, template, method='bilinear') 

# Saving raster 
writeRaster(population_aligned, 'population_aligned.tif', format='GTiff', overwrite=TRUE)

# Identifying high-risk pixels (where suitability raster has value 1)
high_risk_mask <- suitability == 1

# Creating a new raster with population counts in high-risk areas
high_risk_population <- population_aligned * high_risk_mask

# Set low-risk pixels to NA
high_risk_population[high_risk_population == 0] <- NA

# Saving raster 
writeRaster(high_risk_population, 'high_risk_population.tif', format='GTiff', overwrite=TRUE)

# Plot the resulting raster
plot(high_risk_population, main="Population Density in High-Risk Areas", col=terrain.colors(100), na.color="transparent")

# Plotting the resulting raster
tm_shape(high_risk_population) +
  tm_raster(style = "cont", palette = "PuRd", title = "High-Risk Population Density") +
  tm_shape(borno_LGAs) +
  tm_borders(col = "grey") +
  tm_layout(legend.outside = TRUE,
            frame = FALSE,
            legend.title.size = 1,
            legend.text.size = 0.8,
            title.size = 3,
            panel.label.bg.color = "transparent",
            panel.label.color = "black")
```

Now I want to look at the percentage of each LGAs population who are at risk of terrorist attacks so I can consider relative risk across the state.

```{r Percent of population at risk}
# Calculate total population and high-risk population for each LGA
borno_LGAs$total_population <- exact_extract(population_aligned, borno_LGAs, 'sum')
borno_LGAs$high_risk_population <- exact_extract(high_risk_population, borno_LGAs, 'sum')

# Calculate the percentage of population in high-risk areas
borno_LGAs <- borno_LGAs %>%
  mutate(percentage_high_risk = (high_risk_population / total_population) * 100)

# Plot the results
ggplot(borno_LGAs) +
  geom_sf(aes(fill = percentage_high_risk)) +
  scale_fill_viridis_c(name = "Percentage High Risk", 
                       breaks = seq(0, 100, by = 10), 
                       labels = c("0%", "10%", "20%", "30%", "40%", "50%", "60%", "70%", "80%", "90%", "100%"),
                       limits = c(0, 100)) +
  theme_minimal(base_family = "Avenir") +
  labs(title = "Percentage of Population in High-Risk Areas by LGA",
       fill = "Percentage") +
  theme(
    panel.grid = element_blank(),
    axis.title = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank()
  )
```
Finally, I shall do the same but for populations under 15 years old.

```{r Populations under 15 at high risk}
# Re-loading raster
suitability <- raster('suitability_attacks.tif')
child_population <- raster('child_pop.tif')
child_population_aligned <- raster('child_population_aligned.tif')

# Resampling the population raster to match the suitability raster
template <- raster(extent(suitability), res=res(suitability), crs=crs(suitability))
child_population_aligned <- resample(child_population, template, method='bilinear')

# Identifying high-risk pixels (where suitability raster has value 1)
high_risk_mask <- suitability == 1

# Creating a new raster with population counts in high-risk areas
high_risk_child_population <- child_population_aligned * high_risk_mask

# Set low-risk pixels to NA
high_risk_child_population[high_risk_child_population == 0] <- NA

# Saving raster 
writeRaster(high_risk_child_population, 'high_risk_child_population.tif', format='GTiff', overwrite=TRUE)

# Plot the resulting raster
plot(high_risk_child_population, main="Density of population under 15 across high risk areas", col=terrain.colors(100), na.color="transparent")

# Plotting the resulting raster
tm_shape(high_risk_child_population) +
  tm_raster(style = "cont", palette = "PuRd", title = "High-Risk population density, under 15 years old") +
  tm_shape(borno_LGAs) +
  tm_borders(col = "grey") +
  tm_layout(legend.outside = TRUE,
            frame = FALSE,
            legend.title.size = 1,
            legend.text.size = 0.8,
            title.size = 3,
            panel.label.bg.color = "transparent",
            panel.label.color = "black")
```

And now to look at the percentage of the under 15 population per LGA that are at a high risk of terrorist attacks:

```{r Percent of population under 15 at risk}
# Calculate total population and high-risk population for each LGA
borno_LGAs$total_child_population <- exact_extract(child_population_aligned, borno_LGAs, 'sum')
borno_LGAs$high_risk_child_population <- exact_extract(high_risk_child_population, borno_LGAs, 'sum')

# Calculate the percentage of under-15 population in high-risk areas
borno_LGAs <- borno_LGAs %>%
  mutate(percentage_high_risk_child = (high_risk_child_population / total_child_population) * 100)

# Plot the results
ggplot(borno_LGAs) +
  geom_sf(aes(fill = percentage_high_risk_child)) +
  scale_fill_viridis_c(name = "Percentage High Risk", 
                       breaks = seq(0, 100, by = 10), 
                       labels = c("0%", "10%", "20%", "30%", "40%", "50%", "60%", "70%", "80%", "90%", "100%"),
                       limits = c(0, 100)) +
  theme_minimal(base_family = "Avenir") +
  labs(title = "Percentage of Under-15 Population in High-Risk Areas by LGA",
       fill = "Percentage") +
  theme(
    panel.grid = element_blank(),
    axis.title = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank()
  )
```