# Dissertation part 1: Exploratory data analysis

This script contains the exploratory data analysis for my dissertation titled 'A geospatial analysis of terrorism risk across Nigeria'. The aim of this EDA is to look at what kind of terrorism is occurring in Nigeria, where it is occurring and whether any initial patterns/trends can just be visually identified. This will serve as the basis for narrowing down the dissertation research scope (time/space).

## Exploration of terrorism in Nigeria

```{r Loading packages}
library(sf)
library(tmap)
library(ggplot2)
library(dplyr)
library(lubridate)
library(raster)
library(extrafont)
library(spdep)
library(geostan)
library(SpatialEpi)
library(dismo)
library(rJava)
library(terra)
library(pscl)
```

```{r Reading in data}
nigeria <- read_sf('gadm41_NGA_0.shp')
states <- read_sf('gadm41_NGA_1.shp')
LGAs <- read_sf('gadm41_NGA_2.shp')
global_terrorism <- read.csv('global_terrorism.csv', fileEncoding = "latin1")
population <- raster('nga_ppp_2020_1km_Aggregated.tif')
```

Data sources: 
- Shapefile data obtained via GADM: <https://gadm.org> 
- Terrorism data obtained via GTD (requires requesting): <https://www.start.umd.edu/gtd/> 
- Population data obtained via WorldPop (Unconstrained Nigeria 2020 1km): <https://hub.worldpop.org/geodata/listing?id=76>

```{r Cleaning the shapefile datasets for ease of joining later}
# Keeping only necessary columns and doing some renaming for total compatability
states <- dplyr::select(states, 4, 12)
states <- states %>% 
  rename(state = NAME_1)
states <- states %>%
  mutate(state = ifelse(state == "Federal Capital Territory", "Abuja", state))

LGAs <- dplyr::select(LGAs, 5, 7, 14)
LGAs <- LGAs %>% 
  rename(state = NAME_1,
         LGA = NAME_2)
```

### Temporal nature of attacks: Nigeria

```{r Extracting Nigerian terrorism}
nigeria_terrorism <- global_terrorism %>% filter(country_txt == "Nigeria")
```

```{r Cleaning the new Nigerian terrorism dataset}
# Keeping only necessary columns
nigeria_terrorism <- dplyr::select(nigeria_terrorism, 2, 3, 4, 12, 13, 14, 15, 27, 30, 36, 39, 40)

# Removing NAs
nigeria_terrorism <- nigeria_terrorism[!is.na(nigeria_terrorism$latitude) & !is.na(nigeria_terrorism$longitude), ]

# Converting it into a spatial object
nigeria_terrorism <- st_as_sf(nigeria_terrorism, coords = c("longitude", "latitude"), crs = 4326, agr = "constant")

# Plotting terrorist incidents by year
ggplot(nigeria_terrorism, aes(x = iyear)) +
  geom_bar(fill = "#ef6030") +
  labs(title = "Number of Terrorism Incidents by Year",
       x = "Year",
       y = "Count of Incidents") +
  theme_minimal(base_family = "Avenir") +
  theme(panel.grid.major = element_blank(),
        panel.grid.minor = element_blank(),
        text = element_text(family = "Avenir"))
```

Based on this, I can see that terrorism really became an issue from 2012 onwards. However, the Arab Spring begun in 2010 and was a time marked by religious conflict and upheaval. As a result, it makes most sense to analyse geospatial patterns of terrorism from 2010 to 2020 (the end of the dataset), as this accounts for the wider regional conflict that would've undoubtedbly had spillover effects in Nigeria.

```{r Further dataset filtering and cleaning}
# Filtering by the relevant years
recent_terrorism <- nigeria_terrorism %>% filter(iyear >= 2010 & iyear <= 2020)

# Creating a date column
recent_terrorism <- recent_terrorism %>%
  mutate(date = as.Date(paste(iyear, imonth, iday, sep = "-"), format = "%Y-%m-%d"))

# Removing the original separate columns + columns which I later deemed to be unnecessary
recent_terrorism <- recent_terrorism[, -c(1, 2, 3, 9, 10)]

# Renaming columns for clarity
recent_terrorism <- recent_terrorism %>% 
  rename(
    state = provstate, 
    attack_type = attacktype1_txt,
    target_category = targtype1_txt,
  )

# Removing any NAs at last
recent_terrorism <- na.omit(recent_terrorism)

# Saving this as a CSV file so I don't need to run this code every time
write.csv(recent_terrorism, "recent_terrorism.csv", row.names = FALSE)
```

Now I want to have a better look at what kind of attacks are occurring and where they're taking place. This represents the bulk of the EDA.

### Attack types: Nigeria

```{r Kinds of terrorist activities taking place}
# Plotting terrorist activity types
recent_terrorism <- recent_terrorism %>%
  mutate(attack_type = factor(attack_type, levels = names(sort(table(attack_type), decreasing = TRUE))))

# Plotting the bar chart
ggplot(recent_terrorism, aes(x = attack_type)) +
  geom_bar(fill = "#ef6030") +
  labs(title = "Frequency of attack types",
       x = "Attack type",
       y = "Count of Incidents") +
  theme_minimal(base_family = "Avenir") +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(angle = 60, hjust = 1, family = "Avenir"),
    text = element_text(family = "Avenir")
  )
```
The most popular terrorist attack style is armed assault, followed by bombings/explosion and hostage taking.

### Attack targets: Nigeria

```{r Targets of terrorist attacks}
# Plotting terrorist attack targets
recent_terrorism <- recent_terrorism %>%
  mutate(target_category = factor(target_category, levels = names(sort(table(target_category), decreasing = TRUE))))

# Plotting the bar chart
ggplot(recent_terrorism, aes(x = target_category)) +
  geom_bar(fill = "#ef6030") +
  labs(title = "Attack targets",
       x = "Target",
       y = "Count of Incidents") +
  theme_minimal(base_family = "Avenir") +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(angle = 60, hjust = 1, family = "Avenir"),
    text = element_text(family = "Avenir")
  )
```

Private citizens and property are disproportionately targeted by terrorists, suggesting that indeed civilians are disproportionately taking on the brunt of attacks.

### Attack success rate: Nigeria

```{r Success rate of attacks}
# Plotting the success rate of attacks
recent_terrorism$success <- factor(recent_terrorism$success, 
                                   levels = c(0, 1), 
                                   labels = c("Unsuccessful attack", "Successful attack"))

# Create a summary table for the pie chart
success_summary <- as.data.frame(table(recent_terrorism$success))
colnames(success_summary) <- c("AttackStatus", "Count")

# Calculate percentages
success_summary$Percentage <- round(100 * success_summary$Count / sum(success_summary$Count), 1)

# Plotting the pie chart
ggplot(success_summary, aes(x = "", y = Count, fill = AttackStatus)) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar("y") +
  theme_void(base_family = "Avenir") +
  labs(title = "Terrorist attack success rate",
       fill = "Attack status") +
  scale_fill_manual(values = c("Unsuccessful attack" = "#F6DF0A", "Successful attack" = "#ef6030")) +
  geom_text(aes(label = paste0(Percentage, "%")), 
            position = position_stack(vjust = 0.5), 
            family = "Avenir") +
  theme(text = element_text(family = "Avenir"))
```

Of the attacks that are carried out/registered, the vast majority (91%) are successful in causing harm.

### Attack distribution: Nigeria

```{r Plotting attacks as points across Nigeria}
# Plotting points of attacks across the country
ggplot() +
  geom_sf(data = nigeria, fill = NA, color = "black", size = 1.5) + 
  geom_sf(data = states, fill = NA, color = "lightgrey") +
  geom_sf(data = recent_terrorism, color = "#ef6030", size = 0.1) +
  theme_minimal(base_family = "Avenir") +                                       
  labs(title = "2010-2020 Terrorist Attacks in Nigeria") +
  theme(
    panel.grid = element_blank(),        
    axis.title = element_blank(),        
    axis.text = element_blank(),          
    axis.ticks = element_blank(),         
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA),
    text = element_text(family = "Avenir")
  )
```

Clearly, clustering can be seen amongst attacks that have taken place. To understand this a bit better, a thematic map will be generated to identify high-risk states. To ensure that the map is standardised however, terrorist attacks must be plotted as an incident rate.

### Incident rate of attacks: Nigeria

```{r Obtaining the population count for each state}
# Ensuring they have the same crs
states <- st_transform(states, crs(population))

# Extracting the population count per state
population_sum <- function(state, raster) {
  masked_raster <- mask(raster, state)
  sum(getValues(masked_raster), na.rm = TRUE)
}

states$population <- sapply(1:nrow(states), function(i) population_sum(states[i, ], population))

# Creating a data frame that stores state geometries and extracted population data
states_pop <- st_drop_geometry(states) %>%
  mutate(geometry = st_as_text(states$geometry))
```

To create a standarised rate that takes into account population size, I will calculate the rate of terrorist attacks per 10,000 people - a number selected on the basis of relative population size per state.

```{r Calculating the rate of attacks per population counts}
# Creating a data frame of the spatial object (recent_terrorism)
df <- as.data.frame(recent_terrorism)

# Creating a new data frame with the count of occurrences for each state
states_attacks <- df %>%
  group_by(state) %>%
  summarise(count = n()) %>%
  ungroup()

# Creating a new dataframe that contains the population plus terrorist attack count plus geometry
states_terrorism <- states_pop %>%
  left_join(states_attacks, by = "state")

# Calculate attacks per 10,000 people
states_terrorism <- states_terrorism %>%
  mutate(attacks_per_10000 = (count / population) * 10000)

# Convert the geometry column back to sf format if it's not already
states_terrorism <- st_as_sf(states_terrorism, wkt = "geometry")

# Plotting the rate of attacks 
ggplot(states_terrorism) +
  geom_sf(aes(fill = attacks_per_10000)) +
  scale_fill_gradient(low = "white", high = "#ef6030", name = "Attacks per 10,000 people") +
  theme_minimal(base_family = "Avenir") +
  theme(
    panel.grid = element_blank(),      
    axis.title = element_blank(),      
    axis.text = element_blank(),      
    axis.ticks = element_blank()      
  )
```

And to plot this as a bar chart:

```{r Plotting attacks by state}
states_terrorism <- states_terrorism %>%
  mutate(state = factor(state, levels = state[order(-attacks_per_10000)]))

ggplot(states_terrorism, aes(x = state, y = attacks_per_10000)) +
  geom_bar(stat = "identity", fill = "#ef6030") +
  labs(title = "Attack rate by state",
       x = "State",
       y = "Rate per 10,000") +
  theme_minimal(base_family = "Avenir") +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

Evidently, Borno is disproportionately affected by terrorist attacks. I want to now take a closer look at the distribution of terrorist attacks per LGA within the state.

## Terrorism in Borno

```{r Filtering the Borno LGA shapefile}
# Obtaining the extent of Borno from the states shapefile
borno <- states %>% filter(state == "Borno")

# Removing the unnecessary population column 
borno <- borno[, -c(3)]

# Filtering the LGAs shapefile to obtain just those in Borno
borno_LGAs <- LGAs %>% filter(state == "Borno")

# Removing the state column 
borno_LGAs <- borno_LGAs[, -c(1)]
```

```{r Obtaining the population count per LGA in Borno}
# Ensuring the CRS of both the raster and shapefile match
borno <- st_transform(borno, crs(population))

# Converting the shapefile to a spatial object
borno_sp <- as(borno, "Spatial")

# Clipping the raster
population_borno <- mask(population, borno_sp)

# Ensuring the CRS of both the raster and shapefile match
borno_LGAs <- st_transform(borno_LGAs, crs(population_borno))

# Creating a function to calculate population sum for each LGA
calculate_population <- function(lga, raster) {
  # Masking the raster with the LGA geometry
  masked_raster <- mask(raster, lga)
  # Calculating the sum of the population within the LGA
  sum(getValues(masked_raster), na.rm = TRUE)
}

# Applying the function to each LGA
borno_LGAs$population <- sapply(1:nrow(borno_LGAs), function(i) calculate_population(borno_LGAs[i, ], population_borno))

# Making the population count a whole number
borno_LGAs$population <- round(borno_LGAs$population, 0)

# Plotting the population per LGA
ggplot(data = borno_LGAs) +
  geom_sf(aes(fill = population)) +
  scale_fill_gradient(low = "white", high = "#ef6030", name = "Population", 
                      labels = scales::comma) +
  theme_minimal(base_family = "Avenir") +
  labs(title = "Population distribution across Borno",
       fill = "Population") +
  theme(
    panel.grid = element_blank(),      
    axis.title = element_blank(),      
    axis.text = element_blank(),      
    axis.ticks = element_blank()      
  )
```

Finally, I want to look at the distribution of terrorist attacks relative to the population in each LGA.

### Attack distribution: Borno

```{r Plotting terrorist attack points across Borno}
# Obtaining only points in Borno
borno_terrorism <- recent_terrorism %>% filter(state == "Borno")

# Filtering to remove any outliers (ie. points wrongly described as being in Borno)
borno_terrorism <- st_transform(borno_terrorism, crs(borno))

# Perform spatial intersection to filter points within Borno
borno_terrorism <- st_intersection(borno_terrorism, borno)

# Plotting points of attacks across the country
ggplot() +
  geom_sf(data = borno, fill = NA, color = "black", size = 1.5) + 
  geom_sf(data = borno_LGAs, fill = NA, color = "lightgrey") +
  geom_sf(data = borno_terrorism, aes(geometry = geometry), color = "#ef6030", size = 0.3) +
  theme_minimal(base_family = "Avenir") +
  labs(title = "Terrorist Attacks in Borno State, 2010-2020") +
  theme(
    panel.grid = element_blank(),        
    axis.title = element_blank(),        
    axis.text = element_blank(),          
    axis.ticks = element_blank(),         
    plot.background = element_rect(fill = "white", color = NA),
    panel.background = element_rect(fill = "white", color = NA),
    text = element_text(family = "Avenir")
  )
```

### Incident rate of attacks: Borno

```{r Calculating the rate of attacks relative to the population in each LGA}
# Transforming the CRS of borno_terrorism to match borno_LGAs
borno_terrorism <- st_transform(borno_terrorism, crs(borno_LGAs))

# Performing a spatial join to associate each attack with an LGA
attacks_in_LGAs <- st_join(borno_terrorism, borno_LGAs, join = st_within)

# Counting the number of attacks per LGA
attack_counts <- attacks_in_LGAs %>%
  group_by(LGA) %>% 
  summarize(count = n())

# Turning this to a df before joining (dropping the geometry)
attack_counts_df <- st_drop_geometry(attack_counts)

# Joining the new attack_counts_df to borno_LGAs 
borno_LGAs <- borno_LGAs %>%
  left_join(attack_counts_df, by = "LGA") %>%
  rename(attack_count = count)

# Calculating a rate of attacks per 1,000 people
borno_LGAs$attack_per_1000 <- (borno_LGAs$attack_count / borno_LGAs$population) * 1000

# Plotting this as a thematic map
ggplot(borno_LGAs) +
  geom_sf(aes(fill = attack_per_1000)) +
  scale_fill_gradient(low = "white", high = "#ef6030", name = "Attacks per 1,000 people") +
  theme_minimal(base_family = "Avenir") +
  labs(title = "Terrorist attack rate by LGA across Borno state") +
  theme(
    panel.grid = element_blank(),      
    axis.title = element_blank(),      
    axis.text = element_blank(),      
    axis.ticks = element_blank()      
  )
```

And finally, to plot this as a bar chart.

```{r Graphically representing attacks by state}
borno_LGAs <- borno_LGAs %>%
  mutate(LGA = factor(LGA, levels = LGA[order(-attack_per_1000)]))

ggplot(borno_LGAs, aes(x = LGA, y = attack_per_1000)) +
  geom_bar(stat = "identity", fill = "#ef6030") +
  labs(title = "Attack rate by LGA",
       x = "LGA",
       y = "Rate per 1,000") +
  theme_minimal(base_family = "Avenir") +
  theme(
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.text.x = element_text(angle = 45, hjust = 1)
  )
```

It appears that the LGA of Konduga is disproportionately affected by terrorist attacks and that the rate of attacks does not follow an identical pattern to population density.

### Expected risk of terrorist attacks: Borno

I want to create a map that plots the relative risk of attacks, in relation to what is expected.

```{r Calculating the expected number of attacks}
# Calculating the expected number of cases
borno_LGAs$ExpectedNum <- round(expected(population = borno_LGAs$population, cases = borno_LGAs$attack_count, n.strata = 1), 0)

# Converting this to a normalised risk ratio 
borno_LGAs$risk_ratio <- borno_LGAs$attack_count / borno_LGAs$ExpectedNum

max_risk_ratio <- max(borno_LGAs$risk_ratio, na.rm = TRUE)
min_risk_ratio <- min(borno_LGAs$risk_ratio, na.rm = TRUE)

# Rounding this to 2dp for visualisation sake
borno_LGAs$risk_ratio <- round(borno_LGAs$risk_ratio, 2)

# Creating a new column which I can use for colouring my map
borno_LGAs$risk_expectation <- ifelse(borno_LGAs$risk_ratio < 1, "lower", "higher")

borno_LGAs$normalized_risk_ratio <- (borno_LGAs$risk_ratio - min_risk_ratio) / (max_risk_ratio - min_risk_ratio)

# Plotting this
expected_risk <- ggplot(borno_LGAs) +
  geom_sf(aes(fill = risk_expectation), color = "black") +
  scale_fill_manual(
    values = c("higher" = "#FEC4C4", "lower" = "#C4F2FE"),
    name = "Expected risk of attack",
    labels = c("higher" = "Higher than expected", "lower" = "Lower than expected")
  ) +
  geom_text(aes(label = round(risk_ratio, 2), geometry = geometry), 
            stat = "sf_coordinates", size = 3, color = "black", family = "Avenir") +
  theme_minimal(base_family = "Avenir") +
  labs(title = "Expected Risk of Terrorist Attacks in Borno State") +
  theme(
    panel.grid = element_blank(),
    axis.title = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank(),
    text = element_text(family = "Avenir"),
    legend.position = "right"
  )

plot(expected_risk)

ggsave("borno_terrorist_attacks_map.png", plot = expected_risk, width = 15, height = 15, dpi = 300)
```